---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

<<<<<<< HEAD
```{r echo = FALSE, results = 'hide', warning = FALSE, message = FALSE, error = FALSE}
=======
```{r echo = FALSE}
>>>>>>> 6bdc23650ee915093268f2bf2b21cd3bf58d6063
library(ggplot2)
library(lattice)
Sys.setlocale("LC_TIME","English_United States.1252")
```


## Loading and preprocessing the data

```{r}
rawData <- read.csv("activity.csv", sep = ",", header = TRUE)
rawData$date <- as.Date(rawData$date)
hist(rawData$steps, breaks = 20, main="Total number of steps taken per day", xlab="Steps", ylab="Frequency")
```

## What is mean total number of steps taken per day?

```{r}
stepsAggByDate <- aggregate(steps ~ date, data = rawData, sum, na.rm = TRUE)
hist(stepsAggByDate$steps, breaks = 20, main = "Total number of steps taken per day", xlab="Steps", ylab="Frequency")
barplot(stepsAggByDate$steps, names.arg = stepsAggByDate$date)
totalSumSteps <- sum(stepsAggByDate$steps, na.rm = TRUE)
meanSteps <- mean(stepsAggByDate$steps, na.rm = TRUE)
medianSteps <- median(stepsAggByDate$steps, na.rm = TRUE)
```

The total number of steps taken per day is: `r as.integer(totalSumSteps)`

The mean and median values are as following, removing the NAs:  
*Mean*: `r as.integer(meanSteps)`  
*Median*: `r as.integer(medianSteps)`

## What is the average daily activity pattern?

```{r}
avgDailyAct <- aggregate(steps ~ interval, data = rawData, mean, na.rm = TRUE)
plot(avgDailyAct$interval, avgDailyAct$steps, type="l", main = "Average steps per interval (5 min)", xlab = "Interval", ylab = "Steps")
maxStepsX <- which.max(avgDailyAct$steps)
maxStepsY <- avgDailyAct$interval[maxStepsX]
```

5-minute interval *`r as.integer(maxStepsY)`*, on average across all the days in the dataset, contains the maximum number of steps (*= `r as.integer(avgDailyAct$steps[maxStepsX])`*).

## Imputing missing values

```{r}
missingPoints <- sum(is.na(rawData$steps))
numRows <- nrow(rawData)
percentMissing <- missingPoints/numRows * 100
```

There are `r as.integer(missingPoints)` points missing out of `r numRows` records (`r format(round(percentMissing, 2), nsmall = 2)`%).

```{r}
adjustedData <- rawData
stepsMeanByInterval <- aggregate(steps ~ interval, data = rawData, mean, na.rm = TRUE)
adjustedData$steps[is.na(adjustedData$steps)] <- 
  stepsMeanByInterval$steps[match(adjustedData$interval[is.na(adjustedData$steps)], stepsMeanByInterval$interval)]
stepsAggByDateAdj <- aggregate(steps ~ date, data = adjustedData, sum, na.rm = TRUE)
stepsRawAndAdj <- rbind(cbind(stepsAggByDate, mod = "raw"), cbind(stepsAggByDateAdj, mod = "adjusted"))
ggplot(data = stepsRawAndAdj, aes(steps, fill = mod)) + geom_histogram(alpha=.5, bins = 20, position="identity")
ggplot(data = stepsRawAndAdj, aes(x = date, y = steps, fill = mod)) + geom_bar(stat="identity", position = "identity", alpha=.5)
```

The NA values have been replaced with a mean *steps* value from all the same *5-minute intervals*, i.e. if the data is missing for *interval = 5*, we take all the other *steps* values from all the other dates for that *interval* and calculate their mean.

```{r}
totalSumStepsAdj <- sum(stepsAggByDateAdj$steps, na.rm = TRUE)
meanStepsAdj <- mean(stepsAggByDateAdj$steps, na.rm = TRUE)
medianStepsAdj <- median(stepsAggByDateAdj$steps, na.rm = TRUE)
```

The total number of steps taken per day is: `r as.integer(totalSumStepsAdj)`

The mean and median values are as following, removing the NAs:  
*Mean*: `r as.integer(meanStepsAdj)`  
*Median*: `r as.integer(medianStepsAdj)`

While we observe quite a big change in number of steps (`r format(round(totalSumStepsAdj/totalSumSteps*100 - 100, 2), nsmall = 2)`%), the mean and the median are almost identical as for the raw data.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
adjustedData$dayname <- weekdays(adjustedData$date)
adjustedData$weekend <- as.factor(ifelse(adjustedData$dayname == "Saturday" |
                                 adjustedData$dayname == "Sunday", "weekend", "weekday"))
plotdata <- aggregate(steps ~ interval + weekend, data = adjustedData, mean)
xyplot(steps ~ interval | factor(weekend), data = plotdata, aspect = 1/3, type = "l")
```

The above plot contains a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged over all weekdays and weekends. The difference between them is clearly visible.